# High performance computing: data.table {#data.table}

Resources:

- [Vignette Intro](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html)
- [Vignette on reference semantics](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reference-semantics.html)
- [Github Wiki](https://github.com/Rdatatable/data.table/wiki)
- [Cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/master/datatable.pdf)
- [data.table vs. dplyr](https://stackoverflow.com/questions/21435339/data-table-vs-dplyr-can-one-do-something-well-the-other-cant-or-does-poorly)

Suggested data: any

Possible approach: take different sequences of operations in dplyr and translate it into data.table syntax. Then summarise differences and pros/cons vis-à-vis dplyr 

Okay, let's get started with this chapter on high performance computing using the R package data.table.
First we will install exactly this package.

```{r}
print("Hello Data.Table")
install.packages("data.table" , repos="http://cran.us.r-project.org")
```


```{r}
library(tidyverse) 
library(data.table)
inpOlympic <- if (file.exists("data/olympic-games.csv")) {
   "data/olympic-games.csv"
} 


inpWeather <- if (file.exists("data/weather_kiel_holtenau.csv")) {
   "data/weather_kiel_holtenau.csv"
} 

inpSpoti <- if (file.exists("data/spotify_charts_germany.csv")) {
   "data/spotify_charts_germany.csv"
} 

olympics <- fread(inpOlympic)
weather <- fread(inpWeather)
spotify <- fread(inpSpoti)
```
### Data exploration

Datatype data.table

```{r}

class(olympics)
```


## Subsetting rows
We can condition on the rows one time

```{r}
swimmers <- olympics[sport == "Swimming"]
```

and two or more times

```{r}
swimmers <- olympics[sport == "Swimming" & medal == "Gold"]
```

ordering data like

```{r}
var <- swimmers[order(height)]
```

Now we want to condition on the colums too

```{r}
cols <- c("athlete", "medal")
```


or let us use regualr expression to find someone

```{r}
ledecky <- swimmers[athlete %like% "Ledecky"]
```







