# High performance computing: data.table {#data.table}


Okay, let's get started with this chapter on high performance computing using the R package data.table. 

```{r echo=FALSE}
print("Hello data.table")
```

If `data.table` is not yet installed, we install it via:
```{r eval=FALSE}
install.packages("data.table" , repos="http://cran.us.r-project.org")
```

We load `data.table` and - to demonstrate the differences - also the `tidyverse` .
```{r message=FALSE, warning=FALSE}
library(tidyverse) 
library(data.table)
```

Data.table objects are similar to a data.frame object but with some extensions. Let us first load some data with the data.table file reader fread(), which is optimised for performance and creates a data.table object.

```{r}
inpOlympic <- if (file.exists("data/olympic-games.csv")) {
   "data/olympic-games.csv"
} 


inpWeather <- if (file.exists("data/weather_kiel_holtenau.csv")) {
   "data/weather_kiel_holtenau.csv"
} 

inpSpoti <- if (file.exists("data/spotify_charts_germany.csv")) {
   "data/spotify_charts_germany.csv"
} 

olympics <- fread(inpOlympic)
weather <- fread(inpWeather)
spotify <- fread(inpSpoti)
```

## Motivation
The package `data.table` provides an enhanced data frame object called (big surprise: ) **data.table** and a different type of syntax for common types of data manipulations. In this sense it is similar to the tidyverse (especially the dplyr part of it).

So what are the main differences between `data.table` and `tidyverse`? Why or when would we want to use one or the other

- **Performance:** data.tables are heavily optimised for speed. If performance is crucial, then data.tables might be the better choice.
- **Syntax:** data.table syntax is concise while tidyverse syntax is more expressive. Code sequences that span multiple lines are easier to read in case of tidyverse syntax due to the pipe (`%>%`) operator. Overall, the choice is a matter of personal preferences here.
- **Mutability:** data.tables are mutable, i.e. they can be changed in place. If this is a desired feature, then we might opt for data.tables.
- **Quoted and unquoted names:** in data.tables we usually have the option to use either quoted column names (which makes writing functions easier) or unquoted column names (which is convenient for exploration). In dplyr the default is unquoted column names and it is a bit trickier to use quoted column names.

Before we go into the details, let's start with two motivating examples of differences between data.table and tidyverse syntax. First, we want to know the 5 sports with the highest average weight of players in the 2017 summer olympic games.
```{r}
# data.table syntax is short
olympics[game=="2016 Summer", .(weight = mean(weight, na.rm=T)), sport][order(-weight)][1:5]
```

```{r eval=FALSE}
# tidyverse syntax is more expressive and readable
olympics %>% 
  filter(game=="2016 Summer") %>%
  group_by(sport) %>%
  summarise(weight = mean(weight, na.rm=TRUE)) %>%
  arrange(-weight) %>%
  head()
```

Next, we want to change an existing column:
```{r message=FALSE, warning=FALSE}
# data.tables are mutable objects, i.e. changes happen in-place
olympics[,age:=as.numeric(age)]   # age is coerced from character to numeric
```

```{r eval=FALSE}
# tidyverse's tibbles are immutable, i.e. we need to reassign to make permanent changes to the data
olympics <- olympics %>% mutate(age_numeric =as.numeric(age))
```



## Data exploration

The datatype data.table has the following structure: dt[i,j,by], where i conditon on the rows, j select the colums and by defines the groups by which results are aggregated.

```{r}
class(olympics)   # We see that a data.table inherits properties of data.frames
```

### Subsetting rows
We can condition on the rows one time, which gives us all swimmers.

```{r}
class(olympics)
swimmers <- olympics[sport == "Swimming"]
```

and two or more times with the '&' operator ('|' operator for the logical or),for example all gold medal winners in swimming.

```{r}
swimmers_gold <- olympics[sport == "Swimming" & medal == "Gold"]
```

The colums can be treated like variables so the expression 'olympics$sport' is not necessary here but would work too. In additon we can select for the row numbers:

```{r}
var <- swimmers[1:5]
```

This is especially useful after ordering the data:

```{r}
height <- swimmers[order(height)]
```
or in decreasing order
```{r}
height <- swimmers[order(-height)]
```
The order function, called on data.table objects calls the forder() function which is optimised on such objects. Counting rows is possible with different aproaches:

```{r}
usmedals <- olympics[sport == "Swimming" & !is.na(medal), length(medal)]
```
or with the .N approach
```{r}
usmedals <- olympics[sport == "Swimming" & !is.na(medal), .N]
usmedals
```


### Selecting Columns

Now we want to condition on the colums too. Here we can either write the column names as unquoted names inside a list. Note we can use the `.` as an abbreviation for `list` here.  
```{r}
swimmers[, list(athlete, medal)][1:5]   # list with unquoted column names 
swimmers[,    .(athlete, medal)][1:5]   # . is an abbreviation for lists
```
Or we can write the quoted column names inside the `c()` function. 
```{r}
swimmers[,  c("athlete", "medal")][1:5] #
```

This syntax also allows us to define the column in a separate step.
```{r}
cols <- c("athlete", "medal")  # columns are defined in a separate step
swimmers[, ..cols][1:5]        # note the .. syntax  
```
or using the 'with' statement

```{r}
colums <- swimmers[, cols, with = FALSE]
head(colums)
```

and operations on the rows and colums at the same time:


```{r}
var <- olympics[sport == "Swimming" & medal == "Gold", ..cols]
```

### Grouping results

Now we'll consider the third parameter of the data.table object DT[i, j, by].
Let's count the competitions grouped by sports.
(Hint: if we have a look at the data in teamsports, every athlete is listed so caution with this result.)

```{r}
sports <- olympics[, .N, by = "sport"]
```
So using all three arguments we find all swimmers, and grouping they by the country
```{r}
sw <- olympics[sport== "Swimming", .N,by = country]
```
grouping by more colums
```{r}
sw <- olympics[sport== "Swimming", .N,by = .(country,discipline)]
```
If we want to order the data we have to change by to keyby
```{r}
sw <- olympics[sport== "Swimming", .N,keyby = .(country,discipline)]
```
Chaining operations on the data.table object:
```{r}
ordOlym <- olympics[,.N, by = .(country,game)][order(-game,country)]
```


### Editing Data

When we want to edit a Data set of the class dataframe like
```{r}
df = data.frame(name = c("D","a","t","a"), a = 1:4, b = 5:8, c = 9:12)
df
df$c <- 13:16
df
```
this is done via copying the data set and a bad performance. Datatable provides the ':=' operator for an better performance. Here the Data object isn't copied but edited by reference

Adding a column
```{r}
spotify[, ':='(duration_s=duration_ms/1000)]
```
this column is added 'by reference' which includes an higher performance.
Edit an column
```{r}
olympics[sex=='M', sex := 'Male']
olympics[sex=='F', sex := 'Female']
```
Deleting an column by reference
```{r}
spotify[, duration_ms := NULL]
```

## Further resources
Here are links to further resources on the package `data.table`:

- [Vignette Intro](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html)
- [Vignette on reference semantics](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reference-semantics.html)
- [Github Wiki](https://github.com/Rdatatable/data.table/wiki)
- [Cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/master/datatable.pdf)
- [data.table vs. dplyr](https://stackoverflow.com/questions/21435339/data-table-vs-dplyr-can-one-do-something-well-the-other-cant-or-does-poorly)







