# High performance computing: data.table {#data.table}

Resources:

- [Vignette Intro](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html)
- [Vignette on reference semantics](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reference-semantics.html)
- [Github Wiki](https://github.com/Rdatatable/data.table/wiki)
- [Cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/master/datatable.pdf)
- [data.table vs. dplyr](https://stackoverflow.com/questions/21435339/data-table-vs-dplyr-can-one-do-something-well-the-other-cant-or-does-poorly)

Suggested data: any

Possible approach: take different sequences of operations in dplyr and translate it into data.table syntax. Then summarise differences and pros/cons vis-à-vis dplyr 

Okay, let's get started with this chapter on high performance computing using the R package data.table.
First we will install exactly this package.

```{r}
print("Hello Data.Table")
install.packages("data.table" , repos="http://cran.us.r-project.org")
```
We also need to load the data.table library (and the tidyverse as well). Data.table objects are similar to a data.frame object but with some extensions. Let us first load some data with the data.table file reader fread(), which is optimised for performance and creates a data.table object.

```{r}
library(tidyverse) 
library(data.table)
inpOlympic <- if (file.exists("data/olympic-games.csv")) {
   "data/olympic-games.csv"
} 


inpWeather <- if (file.exists("data/weather_kiel_holtenau.csv")) {
   "data/weather_kiel_holtenau.csv"
} 

inpSpoti <- if (file.exists("data/spotify_charts_germany.csv")) {
   "data/spotify_charts_germany.csv"
} 

olympics <- fread(inpOlympic)
weather <- fread(inpWeather)
spotify <- fread(inpSpoti)
```
## Data exploration

The datatype data.table has the following structure: dt[i,j,by], where i conditon on the rows, j select the colums and by aggregates the results.

```{r}
class(olympics)
```

### Subsetting rows
We can condition on the rows one time, which gives us all swimmers.

```{r}
swimmers <- olympics[sport == "Swimming"]
```

and two or more times with the '&' operator ('|' operator for the logical or),for example all gold medal winners in swimming.

```{r}
swimmers_gold <- olympics[sport == "Swimming" & medal == "Gold"]
```

The colums can be treated like variables so the expression 'olympics$sport' is not necessary here but would work too.
In additon we can select for the row numbers:

```{r}
var <- swimmers[1:5]
```

This is especially useful after ordering the data:

```{r}
height <- swimmers[order(height)]
```
or in decreasing order
```{r}
height <- swimmers[order(-height)]
```
The order function, called on data.table objects calls the forder() function which is optimised on such objects.

Counting rows is possible with different aproaches:

```{r}
usmedals <- olympics[sport == "Swimming" & !is.na(medal), length(medal)]
```
or with the .N approach
```{r}
usmedals <- olympics[sport == "Swimming" & !is.na(medal), .N]
usmedals
```


### Selecting Columns

Now we want to condition on the colums too. Here we have the option to store the colum names in an list and use the .. operation:

```{r}
cols <- c("athlete", "medal")
colums <- swimmers[, ..cols]
head(colums)
```
or using the 'with' statement

```{r}
colums <- swimmers[, cols, with = FALSE]
head(colums)
```

and operations on the rows and colums at the same time:


```{r}
var <- olympics[sport == "Swimming" & medal == "Gold", ..cols]
```

### Grouping results

Now we'll consider the third parameter of the data.table object DT[i, j, by].
Let's count the competitions grouped by sports.
(Hint: if we have a look at the data in teamsports, every athlete is listed so caution with this result.)

```{r}
sports <- olympics[, .N, by = "sport"]
```
So using all three arguments we find all swimmers, and grouping they by the country
```{r}
sw <- olympics[sport== "Swimming", .N,by = country]
```
grouping by more colums
```{r}
sw <- olympics[sport== "Swimming", .N,by = .(country,discipline)]
```
If we want to order the data we have to change by to keyby
```{r}
sw <- olympics[sport== "Swimming", .N,keyby = .(country,discipline)]
```
Chaining operations on the data.table object:
```{r}
ordOlym <- olympics[,.N, by = .(country,game)][order(-game,country)]
```


### Editing Data

When we want to edit a Data set of the class dataframe like
```{r}
df = data.frame(name = c("D","a","t","a"), a = 1:4, b = 5:8, c = 9:12)
df
df$c <- 13:16
df
```
this is done via copying the data set and a bad performance. Datatable provides the ':=' operator for an better performance. Here the Data object isn't copied but edited by reference

Adding a column
```{r}
spotify[, ':='(duration_s=duration_ms/1000)]
```
this column is added 'by reference' which includes an higher performance.
Edit an column
```{r}
olympics[sex=='M', sex := 'Male']
olympics[sex=='F', sex := 'Female']
```
Deleting an column by reference
```{r}
spotify[, duration_ms := NULL]
```








