# High performance computing: data.table {#data.table}

Okay, let's get started with this chapter on high performance computing using the R package data.table. 

```{r echo=FALSE}
print("Hello data.table")
```

If `data.table` is not yet installed, we install it via:

```{r eval=FALSE}
install.packages("data.table" , repos="http://cran.us.r-project.org")
```

We load `data.table` package and - to demonstrate the differences - also the `tidyverse` library.

```{r message=FALSE, warning=FALSE}
library(tidyverse) 
library(data.table)
```

`data.table` objects are similar to a `data.frame` object but with some extensions. Let us first load some data with the `data.table` file reader fread(), which is optimised for performance and creates a `data.table` object. For comparison purposes we also load an additional `data.frame`, respectively tibble object. This is necessary because dplyr functions convert `data.table` objects into objects of type `data.frame`.

```{r message=FALSE, warning=FALSE}
olympics <- fread("data/olympic-games.csv")
spotify <- fread("data/spotify_charts_germany.csv")
olympics_df <- read_csv("data/olympic-games.csv")
```

## Motivation
The package `data.table` provides an enhanced data frame object called (big surprise: ) **data.table** and a different type of syntax for common types of data manipulations. In this sense it is similar to the tidyverse (especially the dplyr part of it). A `data.table` object has the following form DT[i,j,by] where i represent the rows, j the columns and by the grouping argument.

```{r}
class(olympics)   # We see that a data.table inherits properties of data.frames
                  # and we can easily cast an data.frame object to data.table

matrix <- data.frame(c(1,2,3),c(4,5,6),c(7,8,9))
class(matrix)
setDT(matrix)
class(matrix)
```

So what are the main differences between `data.table` and `tidyverse`? Why or when would we want to use one or the other

- **Performance:** `data.table` objects are heavily optimised for speed. If performance is crucial, then `data.table` objects might be the better choice.
- **Syntax:** `data.table` syntax is concise while `tidyverse` syntax is more expressive. Code sequences that span multiple lines are easier to read in case of tidyverse syntax due to the pipe (`%>%`) operator. Overall, the choice is a matter of personal preferences here.
- **Mutability:** `data.table` objects are mutable, i.e. they can be changed in place (by reference). If this is a desired feature, then we might opt for `data.table` objects.
- **Quoted and unquoted names:** in `data.table` objects we usually have the option to use either quoted column names (which makes writing functions easier) or unquoted column names (which is convenient for exploration). In dplyr the default is unquoted column names and it is a bit trickier to use quoted column names.

Before we go into the details, let's start with two motivating examples of differences between `data.table` and `tidyverse` syntax. First, we want to know the 5 sports with the highest average weight of players in the 2016 summer olympic games.

```{r}
# data.table syntax is short
olympics[game=="2016 Summer", .(weight = mean(weight, na.rm=T)), sport][order(-weight)][1:5]
```

```{r eval=FALSE}
# tidyverse syntax is more expressive and readable
olympics %>% 
  filter(game=="2016 Summer") %>%
  group_by(sport) %>%
  summarise(weight = mean(weight, na.rm=TRUE)) %>%
  arrange(-weight) %>%
  head()
```

Next, we want to change an existing column:
```{r message=FALSE, warning=FALSE}
# data.tables are mutable objects, i.e. changes happen in-place
olympics[,age:=as.numeric(age)]   # age is coerced from character to numeric
```

```{r eval=FALSE}
# tidyverse's tibbles are immutable, we need to reassign to make permanent changes to the data
olympics_df <- olympics_df %>% mutate(age_numeric =as.numeric(age))
```

## Data exploration

As mentioned before the datatype `data.table` has the following structure: dt[i,j,by], where i sets conditions on the rows, j selects the colums and by defines the groups by which results are aggregated. We'll start by setting conditions on the rows.

### Subsetting rows
We can set conditions on the rows one time, which gives us all swimmers.

Again, for repetition the `tidyverse` syntax:

```{r eval=FALSE}
swimmers <- olympics_df %>% filter(sport == "Swimming")
```
and in `data.table` syntax
```{r}
swimmers <- olympics[sport == "Swimming"]
olympics[sport=="Swimming"]                # or without assigning
```

And few times with the '&' operator ('|' operator for the logical or), for example all gold medal winners in swimming.

```{r}
swimmers_gold <- olympics[sport == "Swimming" & medal == "Gold"]

#the `!` operator negates the logical expression 
swimmers_gold[!country=="Russia" & country!= "China"]  
```

The `i` argument also allows for regular expressions

```{r}
Michael_Phelps <- olympics[sport=="Swimming" & athlete %like% "Phelps" & country %in% c("United States")]
```

The columns can be treated like variables so the expression `olympics$sport` is not necessary here but would work too. In additon we can select for the row numbers:

`tidyverse`:

```{r eval=FALSE}
var <- swimmers %>% slice(1:5)
```

`data.table`:

```{r}
var <- swimmers[1:5]
```

This is especially useful after ordering the data:

```{r}
var <- swimmers[order(height)]
```

or in decreasing order( here we pick the first element, so the tallest swimmer)

```{r}
swimmers[order(-height)][1:1]
```

The order function, called on `data.table` objects calls the internal `forder()` function which is optimised on such objects. Counting rows is possible with different aproaches:

```{r}
medals <- olympics[sport == "Swimming" & !is.na(medal), length(medal)]
```

or with the .N approach:

```{r}
medals <- olympics[sport == "Swimming" & !is.na(medal), .N]
medals
```

As you see above we come now to the second argument of `data.table` objects:

### Selecting Columns

Now we want to set conditions on the colums too. Here we can either write the column names as unquoted names inside a list. Note we can use the `.`-notation as an abbreviation for `list` here.  

```{r}
swimmers_gold[, list(athlete, country, medal)][1:5]   # list with unquoted column names 
swimmers_gold[,    .(athlete,country, medal)][1:5]   # . is an abbreviation for lists
```

Or we can write the quoted column names inside the `c()` function:

```{r}
swimmers[,  c("athlete", "medal")][1:5] #
```

The `tidyverse` syntax for comparison is:

```{r eval=FALSE}
swimmers %>% select("athlete", "medal")
```

This syntax also allows us to define the column in a separate step.

```{r}
cols <- c("athlete", "medal")  # columns are defined in a separate step
swimmers_gold[, ..cols][1:5]        # note the .. syntax  
```

or using the 'with' statement:

```{r}
colums <- swimmers_gold[, cols, with = FALSE]
head(colums)
```

and operations on the rows and colums at the same time:

```{r}
var <- olympics[sport == "Swimming" & medal == "Gold", ..cols]
```

We can also execute functions on the columns e.g. to find out the average age of all olympians

```{r}
olympics[, mean(age, na.rm=T)]
```

and last but not least we can directly plot within the `j` argument (even if the informative value is questionable here)

```{r echo=TRUE, message=FALSE, warning=FALSE}
spotify[,plot(danceability,Streams)]
```

### Grouping results

Now we'll consider the third parameter of the `data.table` object DT[i, j, by].
Let's count the competitions grouped by sports.
(Hint: if we have a look at the data in teamsports, every athlete is listed so caution with this result.)

```{r}
sports <- olympics[, .N, by = "sport"]
sports[1:5]
```

So using all three arguments we find all swimmers, and grouping they by the country:

```{r}
sw <- olympics[sport== "Swimming", .N,by = country]
sw[1:5]
```

grouping by more colums:

```{r}
sw <- olympics[sport== "Swimming", .N,by = .(country,discipline)]
sw[1:5]
```

If we want to order the data we have to change by to keyby.

```{r}
sw <- olympics[sport== "Swimming", .N,keyby = .(country,discipline)]
sw[1:5]
```

Chaining operations on the `data.table` object:

```{r}
ordOlym <- olympics[,.N, by = .(country,game)][order(-game,country)]
ordOlym[1:5]
```

### Editing Data

In contrast to `data.frame` objects, `data.table` objects are mutable. When we want to edit a Data set of the class dataframe like

```{r}
df = data.frame(name = c("D","a","t","a"), a = 1:4, b = 5:8, c = 9:12)
df
df$c <- 13:16
df
```

this is done via copying the data set and results in a bad performance. `Data.table` provides the ':=' operator for an better performance. Here the Data object isn't copied but edited by reference.

Adding a column:

```{r}
spotify[, ':='(duration_s=duration_ms/1000)]
```

this column is added 'by reference' which is done with a higher performance than adding a column to a `data.frame` object. In addition we do not have to assign the expression back to the variable.

Edit an column:

```{r}
olympics[sex=='M', sex := 'Male']
olympics[sex=='F', sex := 'Female']
```

Deleting an column by reference:

```{r}
spotify[, duration_ms := NULL]
```

### Side effects of the mutability

Sometimes we're in the situation that we want to work with a `data.table` object but not on the object itself but on a copy. For example we want to add again the column duration_ms by reference, then the variable spotify_cols change too.

```{r}
spotify_cols <- names(spotify)   # Columnnames
spotify_cols

spotify[, ':='(duration_ms=duration_s*1000)]
spotify_cols
```

If we want to prevent these side effect we have to work on a copy. So again:

```{r}
spotify[, duration_ms := NULL]   # deleting by reference
spotify_cols

spotify_cols <- copy(names(spotify))
spotify[, ':='(duration_ms=duration_s*1000)]
spotify_cols
```

## Runtime comparision

Let us now compare the runtime of some expression made in `data.table` and `tidyverse`. Performance differs depending on the type of operation, and will likely also depend on the size of the data set. Here we test separately sorting, filtering, and grouping/aggregating for our data set of about 270.000 observations. 

```{r eval=FALSE}
install.packages("microbenchmark")
```

```{r paged.print = FALSE}
library(microbenchmark)              # package for performance comparisons
olympics_tbl <- as_tibble(olympics)  # create a tibble to compare performance

# Sorting
microbenchmark(
  olympics_tbl %>% arrange(age),     # dplyr
  olympics[order(age)],              # data.table
  times = 5                          # run 5 time to get more robust measurements
) 

# Filtering
microbenchmark(
  olympics_tbl %>% filter(height > 160),   # dplyr
  olympics[height > 160],                  # data.table
  times = 5
)

# Grouping
microbenchmark(
  olympics_tbl %>% group_by(athlete) %>% summarise(mean(height, na.rm=T)),  # dplyr
  olympics[, .(mean(height, na.rm=T)), keyby=athlete],                      # data.table
  times = 5
) 
```

## Further resources
Here are links to further resources on the package `data.table`:

- [Vignette Intro](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html)
- [Vignette on reference semantics](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reference-semantics.html)
- [Github Wiki](https://github.com/Rdatatable/data.table/wiki)
- [Cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/master/datatable.pdf)
- [data.table vs. dplyr](https://stackoverflow.com/questions/21435339/data-table-vs-dplyr-can-one-do-something-well-the-other-cant-or-does-poorly)







