# Categorical data: forcats {#forcats}
## Introduction
This chapter is dedicated to the handling of categorical variables. This becomes important if information is to be presented in a non-alphabetical order or aggregated in a meaningful way. Within the R programming language, categorical variables are converted into a form that can be used for analysis using factors. While many Base-R functions automatically convert character vectors into factors, tidyverse requires an explicit treatment. The core-tidyverse provides the package forcats, which will be described here. Further information and exercises are available at the sources shown.

Resources:

- [Homepage](https://forcats.tidyverse.org/)
- [Cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/factors.pdf)
- [Chapter in R4DS](https://r4ds.had.co.nz/factors.html)
- [Vignette](https://forcats.tidyverse.org/articles/forcats.html)

In this chapter the demo data set "diamonds" from the ggplot2-package ([more information](https://ggplot2.tidyverse.org/reference/diamonds.html)) and a dataset "olympic-games" which shows the medal success of olympic athletes from 1896 to 2016 are used. Before you start reading, you should have read the chapter ***Wrangling Data: dplyr*** or already have basic knowledge in this field. 

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
library(tidyverse)                                     # the package tidyverse contains all libraries necessary for this chapter
diamonds <- read_csv("data/diamonds.csv")              # Importing the "diamonds" dataset
olympic <- read_csv("data/olympic-games.csv")          # Importing the "olympic-games" dataset
head(diamonds, 2)                                      # Have a short look at the data
head(olympic,2 )                                       # Have a short look at the data
```

## General functions
### Create
Basically two things are needed to create a factor. A vector which contains the values to be analyzed and a vector which divides the values into levels. As an example we will use the column "clarity" of the diamonds dataset. It is a categorical evaluation of the clarity of a diamond (I1 (worst), SI2, SI1, VS2, VS1, VVS2, VVS1, IF (best)). If you are interested in the distribution of the diamonds in this category, you could do this by using a suitable query:

```{r}
diamonds %>%
  group_by(clarity) %>%
  count() %>%
  ggplot(aes(clarity, n)) +
  geom_col()
```

A sorting of the x-axis, which is based on the degree of clarity, is unfortunately not possible in this form. A workaround is to convert this column into a factor, which allows us to evaluate the individual categories. For this purpose we first define a vector, which ranks the categories according to their value:

```{r}
levels_clarity <- c("I1", "SI2", "SI1", "VS2", "VS1", "VVS2", "VVS1", "IF")
```

In the next step we lay this newly created levels vector on the column "clarity" of our data set to create a factor. ***factor()*** is a base-R function.

```{r}
factor_clarity <- factor(x = diamonds$clarity, levels = levels_clarity)
```

Another possibility to create a factor is to convert a vector using ***as_factor().***. Here the levels are created automatically. Their order depends on the appearance of the corresponding value in the source vector.

```{r}
as_factor(diamonds$clarity) %>%
  levels()
```


### Count values per level
Now we repeat the analysis of the distribution within our dataset using the created factor. The function ***fct_count()*** returns the frequency of a categorical value within a factor. The order of the levels remains unchanged.

```{r}
fct_count(factor_clarity) %>%
  ggplot(aes(clarity, n)) +          #the definition of aes. mapping in this line is only used to label the axes
  geom_col(aes(f, n))
```

It becomes obvious that the distribution can now be displayed in the desired order (by valence). Functions of the package forcats always start with the prefix ***fct_***. 

### Inspect and set levels
With the function ***levels()*** the levels of a factor can be both read and defined. **But be aware **: this can only be used to change the names, not the order of the levels. The base-R function ***unclass()*** gives information about the internal memory structure of a factor.

```{r}
factor_clarity %>% levels()                   #shows the original levels of our factor
factor_clarity[1:25] %>% unclass()            #shows the internal structure
#remain the semantic order but replace the technical jargon with something understandable
levels_clarity_c <- c("c8", "c7", "c6", "c5", "c4", "c3", "c2", "c1")
levels(factor_clarity) <- levels_clarity_c   
factor_clarity %>% head(25)
```

### Inspect unique values 
The function ***fct_unique*** can be used to output unique values of a factor. In contrast to the base-R function ***unique()*** the values are returned in the order of the levels and not in the order of their appearance.

```{r}
factor_clarity %>% fct_unique()
factor_clarity %>% unique()
```

## Combine factors
### Combine factors with different levels
With the function ***fct_c()*** factors with different levels can be combined to one factor covering all levels. First we create two factors. The first one shows the amount of contested disciplines by several countries during the olympic sommer games 1896. The second one shows the amount of contested disciplines by countries during the olympic winter games 1924.

```{r}
olympic_1896 <- olympic %>% filter(game == "1896 Summer")
summer_1896 <- factor(olympic_1896$country)
olympic_1924 <- olympic %>% filter(game == "1924 Winter")
winter_1924 <- factor(olympic_1924$country)
```

The factors differ both in their stored categorical data and in their set levels. A plot makes this clear:

```{r}
summer_1896 %>%
  fct_count() %>%
  ggplot(aes(`amount of disciplines`, country))+
  geom_col(aes(n,f))

winter_1924 %>%
  fct_count() %>%
  ggplot(aes(`amount of disciplines`, country))+
  geom_col(aes(n,f))
```

Now we combine the factors with ***fct_c()*** and plot it:

```{r}
fct_c(summer_1896, winter_1924) %>%
  fct_count() %>%
  ggplot(aes(`amount of disciplines`, country))+
  geom_col(aes(n,f))
```

Both the underlying categorical data and the defined levels were combined in one factor.

### Standardise levels of various factors
With the function ***fct_unify*** the levels of different factors can be standardised. Please note, both input and output are in list form.
```{r}
factor_list <- fct_unify(list(summer_1896, winter_1924))

# plotting the first factor
factor_list[[1]] %>%
  fct_count() %>%
  ggplot() + 
  geom_col(aes(n, f))

# plotting the second factor
factor_list[[2]] %>%
  fct_count() %>%
  ggplot() + 
  geom_col(aes(n, f))
```

In this case, the underlying categorical data was left unchanged for both factors, but the levels were standardised. This is especially useful when comparing the categorical data of two different factors. 

## Order of levels
### Manual reordering of levels
With the function ***fct_relevel*** the levels of a factor can be reordered. In contrast to the function ***levels()***, which only allows the renaming of factor levels, ***fct_relevel*** also adjusts the internal memory structure of the factor. An example should clarify this.

```{r}
# have a look at the original levels
factor_clarity %>% 
  fct_count()
# The wrong approach (this is semantical wrong):
levels(factor_clarity) <- c("c1","c3","c5","c7","c2","c4","c6","c8")
factor_clarity %>% 
  fct_count()
#tidy up
levels(factor_clarity) <- levels_clarity_c
# The right approach:
factor_clarity %>%
  fct_relevel(c("c1","c3","c5","c7","c2","c4","c6","c8")) %>%
  fct_count()
```

As you can see, only the function ***fct_relevel*** allows a correct relevel.

### Reordering by frequency
Especially for plots it is often useful to orient the order of the levels on the frequency of the corresponding values. The function ***fct_infreq()*** allows exactly this. Plotting the unsorted factor leads to a difficult readability.

```{r}
summer_1896 %>%
  fct_count() %>%
  ggplot(aes(n,f)) +
  geom_col()
```

The better approach is to sort the data before plotting.

```{r}
summer_1896 %>%
  fct_infreq() %>%    #insert fct_infreg() to get the data ordered
  fct_count() %>%
  ggplot(aes(n,f)) +
  geom_col()
```

### Reordering by appearance 
The package ***forcats*** offers with the function ***fct_inorder()*** the possibility to orientate the order of the levels of a factor to the occurence of the corresponding categorical data in the data set. A look at the first 20 entries in the corresponding data sets makes this clear:

```{r}
olympic_1896$country %>%
  head(20)
```

Create a factor in the appropriate order of levels:

```{r}
olympic_1896$country %>%
  fct_inorder() %>%
  levels()
```


## Value of levels
## Add or drop levels
