# Categorical data: forcats {#forcats}
## Introduction
This chapter is dedicated to the handling of categorical variables. This becomes important if information is to be presented in a non-alphabetical order or aggregated in a meaningful way. Within the R programming language, categorical variables are converted into a form that can be used for analysis using factors. While many Base-R functions automatically convert character vectors into factors, tidyverse requires an explicit treatment. The core-tidyverse provides the package forcats, which will be described here. Further information and exercises are available at the sources shown.

Resources:

- [Homepage](https://forcats.tidyverse.org/)
- [Cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/factors.pdf)
- [Chapter in R4DS](https://r4ds.had.co.nz/factors.html)
- [Vignette](https://forcats.tidyverse.org/articles/forcats.html)

In this chapter we use the demo data set "diamonds" from the ggplot2-package ([more information](https://ggplot2.tidyverse.org/reference/diamonds.html)) as well as a dataset "olympic-games" which shows the medal success of olympic athletes from 1896 to 2016. For the latter we focus on the summer games 1896 and the winter games 1924 for practical reasons. Before you start reading, you should have read the chapter ***Wrangling Data: dplyr*** or  be familiar with this field. 

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
library(tidyverse)                                     # the package tidyverse contains all libraries necessary for this chapter
diamonds <- read_csv("data/diamonds.csv")              # Importing the "diamonds" dataset
olympic <- read_csv("data/olympic-games.csv")          # Importing the "olympic-games" dataset
head(diamonds, 2)                                      # Have a short look at the data
head(olympic, 2)                                       # Have a short look at the data
```

## General functions
### Create
Basically two things are needed to create a factor:

1. a vector which contains the values to be analyzed
2. another vector which divides the values into levels

As an example we will use the column `clarity` of the diamonds dataset.
It is a categorical evaluation of the clarity of a diamond and a subset of the grades according to the Gemological Institute of America (GIA) grading system. The grades according to GIA read as follows:

|GIA grade|Category                      |Included in `diamonds`             |
|---------|------------------------------|:---------------------------------:|
|FL       |Flawless                      |<font color="red">&#10008;</font>  | 
|IF       |Internally Flawless           |<font color="green">&#10004;</font>|
|VVS~1~   |Very Very Slightly Included~1~|<font color="green">&#10004;</font>|
|VVS~2~   |Very Very Slightly Included~2~|<font color="green">&#10004;</font>|
|VS~1~    |Very Slightly Included~1~     |<font color="green">&#10004;</font>|
|VS~2~    |Very Slightly Included~2~     |<font color="green">&#10004;</font>|
|SI~1~    |Slightly Included~1~          |<font color="green">&#10004;</font>|
|SI~2~    |Slightly Included~2~          |<font color="green">&#10004;</font>|
|I~1~     |Included~1~                   |<font color="green">&#10004;</font>|
|I~2~     |Included~2~                   |<font color="red">&#10008;</font>  |
|I~3~     |Included~3~                   |<font color="red">&#10008;</font>  |

If you are interested in the distribution of the diamonds in this category, you could do this by using a suitable query:

```{r}
diamonds %>%
  group_by(clarity) %>%
  count() %>%              # count the number of observations per clarity grade
  ggplot(aes(clarity, n)) +
  geom_col()
```

A sorting of the x-axis, which follows the order of the grades as shown in the table above, is unfortunately not possible in this form. A workaround is to convert the column `clarity` into a factor, which allows us to evaluate the individual categories. For this purpose we first define a vector, which ranks the categories according to their grade (from *bad* to *good*):

```{r}
levels_clarity <- c("I1", "SI2", "SI1", "VS2", "VS1", "VVS2", "VVS1", "IF")
```

In the next step we lay this newly created levels vector on the column `clarity` of our data set to create a factor. `factor()` is a base-R function.

```{r}
factor_clarity <- factor(x = diamonds$clarity, levels = levels_clarity)
```
`factor_clarity` consists solely of the `clarity` column of the original `diamonds` data set.

Another possibility to create a factor is to convert a vector using `as_factor()`. Here the levels are created automatically. Their order however depends on the appearance of the corresponding value in the source vector.

```{r}
as_factor(diamonds$clarity) %>%
  levels()
```


### Count values per level
Now we repeat the analysis of the distribution within our dataset using the created factor. The function `fct_count()` returns the frequency of a categorical value within a factor. The order of the levels remains unchanged.

```{r}
fct_count(factor_clarity) %>%
  ggplot(aes(clarity, n)) +          #the definition of aes. mapping in this line is only used to label the axes
  geom_col(aes(f, n))
```

It becomes obvious that the distribution can now be displayed in the desired order (by valence). Functions of the package forcats always start with the prefix `fct_`. 

### Inspect and set levels {#anchor}
With the function `levels()` the levels of a factor can be both read and defined. **But be aware:** this can only be used to change the names, not the order of the levels. The base-R function `unclass()` gives information about the internal memory structure of a factor.

```{r}
factor_clarity %>% levels()                   #shows the original levels of our factor
factor_clarity[1:25] %>% unclass()            #shows the internal structure
#remain the semantic order but replace the technical jargon with something understandable
levels_clarity_c <- c("c8", "c7", "c6", "c5", "c4", "c3", "c2", "c1")
levels(factor_clarity) <- levels_clarity_c   
factor_clarity %>% head(25)
```

### Inspect unique values 
The function `fct_unique()` can be used to output unique values of a factor. In contrast to the base-R function `unique()` the values are returned in the order of the levels and not in the order of their appearance.

```{r}
factor_clarity %>% fct_unique()
factor_clarity %>% unique()
```

## Combine factors
### Combine factors with different levels
With the function `fct_c()` factors with different levels can be combined to one factor covering all levels. First we create two factors. The first one shows the amount of contested disciplines by several countries during the olympic sommer games 1896. The second one shows the amount of contested disciplines by countries during the olympic winter games 1924.
Note, that the default option for the `levels` argument in `factor()` is a sorted set of the given values. As R sorts characters lexicographically. the corresponding levels are the countries sorted alphabetically from A to Z.
```{r}
olympic_1896 <- olympic %>% filter(game == "1896 Summer")
# Using default option for `levels` argument in factor()
countries_in_1896 <- factor(olympic_1896$country)

olympic_1924 <- olympic %>% filter(game == "1924 Winter")
# The default option for the `levels` argument in factor
countries_in_1924 <- factor(olympic_1924$country)
```

The factors differ both in their stored categorical data and in their set levels. A plot makes this clear:

```{r}
countries_in_1896 %>%
  fct_count() %>%
  ggplot(aes(`amount of disciplines`, country))+
  geom_col(aes(n, f))

countries_in_1924 %>%
  fct_count() %>%
  ggplot(aes(`amount of disciplines`, country))+
  geom_col(aes(n, f))
```

Now we combine the factors with `fct_c()` and plot it:

```{r}
fct_c(countries_in_1896, countries_in_1924) %>%
  fct_count() %>%
  ggplot(aes(`amount of disciplines`, country))+
  geom_col(aes(n,f))
```

Both the underlying categorical data and the defined levels were combined in one factor.

### Standardise levels of various factors
With the function `fct_unify()` the levels of different factors can be standardised. Please note, both input and output are in list form.
```{r}
factor_list <- fct_unify(list(countries_in_1896, countries_in_1924))

# plotting the first factor
factor_list[[1]] %>%
  fct_count() %>%
  ggplot() + 
  geom_col(aes(n, f))

# plotting the second factor
factor_list[[2]] %>%
  fct_count() %>%
  ggplot() + 
  geom_col(aes(n, f))
```

In this case, the underlying categorical data was left unchanged for both factors, but the levels were standardised. This is especially useful when comparing the categorical data of two different factors. 

## Order of levels
### Manual reordering of levels
With the function `fct_relevel()` the levels of a factor can be reordered. In contrast to the function `levels()`, which only allows the renaming of factor levels, `fct_relevel()` also adjusts the order of the levels themselves, that is the way they are stored internally. An example should clarify this.

```{r}
# have a look at the original levels
factor_clarity %>% 
  fct_count()
# The wrong approach (this is semantical wrong):
levels(factor_clarity) <- c("c1","c3","c5","c7","c2","c4","c6","c8")
factor_clarity %>% 
  fct_count()
#tidy up
levels(factor_clarity) <- levels_clarity_c
# The right approach:
factor_clarity %>%
  fct_relevel(c("c1","c3","c5","c7","c2","c4","c6","c8")) %>%
  fct_count()
```

As you can see, only the function `fct_relevel()` allows a correct relevel.

### Reordering by frequency
Especially for plots it is often useful to orient the order of the levels on the frequency of the corresponding values. The function `fct_infreq()` allows exactly this. Plotting the unsorted factor leads to a difficult readability.

```{r}
countries_in_1896 %>%
  fct_count() %>%
  ggplot(aes(n, f)) +
  geom_col()
```

The better approach is to sort the data before plotting.

```{r}
countries_in_1896 %>%
  fct_infreq() %>%    #insert fct_infreq() to get the data ordered
  fct_count() %>%
  ggplot(aes(n, f)) +
  geom_col()
```

### Reordering by appearance 
The package `forcats` offers with the function `fct_inorder()` the possibility to orientate the order of the levels of a factor to the occurence of the corresponding categorical data in the data set. To make this clear, we take a look at the unique values with the first 20 entries in the corresponding data set:

```{r}
olympic_1896$country %>%
  head(20) %>%
  unique()
```

Create a factor in the appropriate order of levels:

```{r}
olympic_1896$country %>%
  fct_inorder() %>%
  levels()
```

### Reverse level order
The function `fct_rev()` reverses the existing order of the levels of a factor. First have a look at the original order:
```{r}
countries_in_1896 %>%
  fct_infreq() %>%    #insert fct_infreq() to get the data ordered
  fct_count() %>%
  ggplot(aes(n, f)) +
  geom_col()
```
Now we reverse the order:

```{r}
countries_in_1896 %>%
  fct_infreq() %>%    #insert fct_infreq() to get the data ordered
  fct_rev() %>%       #insert fct_rev() to reverse this order
  fct_count() %>%
  ggplot(aes(n, f)) +
  geom_col()
```

### Shift levels
The argument `n` in the function `fct_shift()` allows to shift the levels to the left (right) for negative (positive) integer values of `n`, wrapping around end.
Thus a value of `n = -1L` would shift the order of the levels to the left by one location.

```{r}
countries_in_1896 %>%
  fct_infreq() %>%          #insert fct_infreq() to get the data ordered
  fct_shift(n = -1L) %>%    #insert fct_shift() to shift levels
  fct_count() %>%
  ggplot(aes(n, f)) +
  geom_col()
```

### Randomly permute levels
The level of a factor can also be randomly shuffeled using `fct_shuffle()`. The input argument can be either a factor or a character vector, whereas the output will be a factor. By way of example, this is demonstrated using the factor `countries_in_1896`:
```{r}
orig_country_levels <- countries_in_1896 %>% levels() # the original levels are sorted alphabetically
orig_country_levels %>% print()

shuffled_country_factor <- orig_country_levels %>% fct_shuffle()
shuffled_country_levels <- shuffled_country_factor %>% levels() # the shuffled levels are randomly sorted
shuffled_country_levels %>% print()
```
### Reordering levels by other variables
The functions presented in this section bare great similarity to the `fct_relevel()` function introduced in the beginning of this section.
`fct_relevel()` allows for a direct manipuation of the levels by passing the new order to the `levels` keyword argument. In this regard `fct_reorder()` and `fct_reorder2()` are different, as the levels are reordered according to the result of a function applied to one vector `x` in case of `fct_reorder()` and two vectors `x` and `y` in case of `fct_reorder2()`.

**Note:** We want to point out a few things that have to be kept in mind when using these functions

: 1. The length of the factor `f` to be reordered has the be equal to the lenght of the vector(s) passed to the function. 
That is an error will be provoked, if `length(f) != length(x)` or `length(f) != length(y)`.
2. In case the function returns the same value for two different elements of the vector(s), R will arrange the levels automatically.
3. In case there are multiple occurences of an element, the level assigned to this element will be based on the result of the function for the first corresponding elements in `x` (and `y`).

The example below is intended to unravel this behavior.
```{r}
# TODO: Need to think of an example here. 
# fct_reorder() for fun(x); fct_reorder2() for fun(x, y)
```

## Change the value of levels
### Renaming the levels
Let's say you want to change the name of the levels (which also impies changing the corresponding value elements) because you are unhappy with the current naming. `fct_recode()` allows to manually give new names to certain levels, without affecting the order of the levels or levels that are not included in the function call. 
The `diamonds$clarity` column is an ideal example for the use of `fct_recode()`. For you conviniece we repeat some of the steps made in the beginning of this chapter:
```{r}
# Define the correct order of the levels in ascending order
levels_clarity <- c("I1", "SI2", "SI1", "VS2", "VS1", "VVS2", "VVS1", "IF")
# Make `diamonds$clarity` a factor and assign the correctly ordered levels
factor_clarity <- factor(x = diamonds$clarity, levels = levels_clarity)

# Since the level notation is somewhat cryptic, we want to change it
renamed_factor_clarity <- factor_clarity %>% 
  fct_recode(`Included 1 (worst)`="I1",
             `Slightly Included 2`="SI2",
             `Slightly Included 1`="SI1",
             `Very Slightly Included 2`="VS2",
             `Very Slightly Included 1`="VS1",
             `Very Very Slightly Included 2`="VVS2",
             `Very Very Slightly Included 1`="VVS1",
             `Internally Flawless (best)`="IF")

```
Note, that this approach is in fact similar to the steps described in  [5.2.3  Inspect and set levels](#anchor).

Another scenario might be that we want to either a pre- or a suffix or both to the current level. To this end, the forcats package provides the `fct_relabel()` function. Its second argument (or its first argument in case of piping) is a function that has to return a character vector. The `countries_in_1896` factor variable is used to showcase how `fct_relabel()` works.
```{r}
add_prefix <- function(input_char) {
  paste("Country:", input_char)
}
# the `add_prefix` function is applied to each level returning a character vector. 
prefixed_countries_in_1896 <- countries_in_1896 %>% fct_relabel(add_prefix)

prefixed_countries_in_1896 %>% levels()
```


## Add or drop levels
