# Categorical data: forcats {#forcats}

## Introduction
This chapter is dedicated to the handling of categorical variables. This becomes important if information is to be presented in a non-alphabetical order or aggregated in a meaningful way. Within the R programming language, categorical variables are converted into a form that can be used for analysis using factors. While many Base-R functions automatically convert character vectors into factors, tidyverse requires an explicit treatment. The core-tidyverse provides the package forcats, which will be described here. Further information and exercises are available at the sources shown.

Resources:

- [Homepage](https://forcats.tidyverse.org/)
- [Cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/factors.pdf)
- [Chapter in R4DS](https://r4ds.had.co.nz/factors.html)
- [Vignette](https://forcats.tidyverse.org/articles/forcats.html)

In this chapter the demo data set "diamonds" from the ggplot2-package ([more information](https://ggplot2.tidyverse.org/reference/diamonds.html)) and a dataset "olympic-games" which shows the medal success of olympic athletes from 1896 to 2016 are used.

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
library(tidyverse)
diamonds<-read_csv("data/diamonds.csv")
olympic<-read_csv("data/olympic-games.csv")

head(diamonds)
head(olympic)
```

## General functions
### Create
Basically two things are needed to create a factor. A vector which contains the values to be analyzed and a vector which divides the values into levels. As an example we will use the column "clarity" of the diamonds dataset. It is a categorical evaluation of the clarity of a diamond (I1 (worst), SI2, SI1, VS2, VS1, VVS2, VVS1, IF (best)). If you are interested in the distribution of the diamonds in this category, you could do this by using a suitable query:

```{r}
diamonds%>%
  group_by(clarity)%>%
  count()%>%
  ggplot(aes(clarity, n))+
  geom_col()
```

A sorting of the x-axis, which is based on the degree of clarity, is unfortunately not possible in this form. A workaround is to convert this column into a factor, which allows us to evaluate the individual categories. For this purpose we first define a vector, which ranks the categories according to their value:

```{r}
levels_clarity<-c("I1", "SI2", "SI1", "VS2", "VS1", "VVS2", "VVS1", "IF")
```

In the next step we lay this newly created levels vector on the column "clarity" of our data set to create a factor. This is a base-R function.

```{r}
factor_clarity<-factor(diamonds$clarity, levels_clarity)
```

### Count values per level
Now we repeat the analysis of the distribution within our dataset using the created factor.

```{r}
fct_count(factor_clarity)%>%
  ggplot(aes(clarity, n))+          #the definition aes is only used here to label the axes
  geom_col(aes(f, n))
```
It becomes obvious that the distribution can now be displayed in the desired order (by valence). Functions of the package forcats always start with the prefix "fct_".

### Inspect the levels



### Inspect unique values 
## Combine factors
## Order of levels
## Value of levels
## Add or drop levels
