# Categorical data: forcats {#forcats}
## Introduction
This chapter is dedicated to the handling of categorical variables. This becomes important if information is to be presented in a non-alphabetical order or aggregated in a meaningful way. Within the R programming language, categorical variables are converted into a form that can be used for analysis using factors. While many Base-R functions automatically convert character vectors into factors, tidyverse requires an explicit treatment. The core-tidyverse provides the package forcats, which will be described here. Further information and exercises are available at the sources shown.

Resources:

- [Homepage](https://forcats.tidyverse.org/)
- [Cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/factors.pdf)
- [Chapter in R4DS](https://r4ds.had.co.nz/factors.html)
- [Vignette](https://forcats.tidyverse.org/articles/forcats.html)

In this chapter we use the demo data set `diamonds` from the ggplot2-package ([more information](https://ggplot2.tidyverse.org/reference/diamonds.html)) as well as a dataset `olympic-games` which shows the medal success of olympic athletes from 1896 to 2016. For the latter we focus on the summer games 1896 and the winter games 1924 for practical reasons.  
Before you start reading, you should have read the chapter ***Wrangling Data: dplyr*** or  be familiar with this field. 

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
# The package tidyverse contains all libraries necessary for this chapter.
library(tidyverse)

# Importing the "diamonds" dataset.
diamonds <- read_csv("data/diamonds.csv")
# Importing the "olympic-games" dataset.
olympic <- read_csv("data/olympic-games.csv")

# Have a short look at the data.
head(diamonds, 2)
# Have a short look at the data.
head(olympic, 2)              
```

## General functions
### Create
Basically two things are needed to create a factor:

1. a vector which contains the **values** to be analyzed
2. another vector which divides the values into **levels**

As an example we will use the column `clarity` of the diamonds dataset.
It is a categorical evaluation of the clarity of a diamond and a subset of the grades according to the Gemological Institute of America (GIA) grading system. The grades according to GIA read as follows:

|GIA grade|Category                      |Included in `diamonds`             |
|---------|------------------------------|:---------------------------------:|
|FL       |Flawless                      |<font color="red">&#10008;</font>  | 
|IF       |Internally Flawless           |<font color="green">&#10004;</font>|
|VVS~1~   |Very Very Slightly Included~1~|<font color="green">&#10004;</font>|
|VVS~2~   |Very Very Slightly Included~2~|<font color="green">&#10004;</font>|
|VS~1~    |Very Slightly Included~1~     |<font color="green">&#10004;</font>|
|VS~2~    |Very Slightly Included~2~     |<font color="green">&#10004;</font>|
|SI~1~    |Slightly Included~1~          |<font color="green">&#10004;</font>|
|SI~2~    |Slightly Included~2~          |<font color="green">&#10004;</font>|
|I~1~     |Included~1~                   |<font color="green">&#10004;</font>|
|I~2~     |Included~2~                   |<font color="red">&#10008;</font>  |
|I~3~     |Included~3~                   |<font color="red">&#10008;</font>  |

If you are interested in the distribution of the diamonds in this category, you could do this by using a suitable query:

```{r}
# Count the number of observations per clarity grade and plot the result.
diamonds %>%
  group_by(clarity) %>%
  count() %>%   
  ggplot(aes(x = n, y = clarity)) +
  geom_col()
```

A sorting of the x-axis, which follows the order of the grades as shown in the table above, is unfortunately not possible in this form. A workaround is to convert the column `clarity` into a factor, which allows us to evaluate the individual categories. For this purpose we first define a vector, which ranks the categories according to their grade (from *bad* to *good*):

```{r}
levels_clarity <- c("I1", "SI2", "SI1", "VS2", "VS1", "VVS2", "VVS1", "IF")
```

In the next step we lay this newly created levels vector on the column `clarity` of our data set to create a factor. The function `factor()` is a base-R function.

```{r}
factor_clarity <- factor(x = diamonds$clarity, levels = levels_clarity)
```
The factor `factor_clarity` consists solely of the `clarity` column of the original `diamonds` data set.

Another possibility to create a factor is to convert a vector using `as_factor()`. Here the levels are created automatically. Their order however depends on the appearance of the corresponding value in the source vector.

```{r}
as_factor(diamonds$clarity) %>%
  levels()
```


### Count values per level
Now we repeat the analysis of the distribution within our dataset using the created factor with the levels in correct order. The function `fct_count()` returns the frequency of a categorical value within a factor. The order of the levels remains unchanged.
```{r}
fct_count(factor_clarity) %>%
  # The definition of aesthetic mapping in this line is only used to label the axes.
  ggplot(aes(x = n, y = clarity)) +          
  geom_col(aes(x = n, y = f))
```

It becomes obvious that the distribution can now be displayed in the desired order (by order of levels).  Functions of the package forcats always start with the prefix `fct_`. 

### <a name="Inspect-and-set-levels"></a>Inspect and set levels
With the function `levels()` the levels of a factor can be both read and renamed. **But be aware:** this can only be used to change the names, not the order of the levels. The base-R function `unclass()` gives information about the internal memory structure of a factor.

```{r}
# Shows the original levels of our factor.
factor_clarity %>% 
  levels()

# Shows the internal structure.
factor_clarity[1:25] %>% 
  unclass()               

# Keep the order of levels,
# but replace the technical GIA jargon with something more understandable.
levels_clarity_c <- c("c8", "c7", "c6", "c5", "c4", "c3", "c2", "c1")
levels(factor_clarity) <- levels_clarity_c

factor_clarity %>% 
  head(25)
```

### Inspect unique values 
The function `fct_unique()` can be used to output unique values of a factor. In contrast to the base-R function `unique()` the values are returned in the order of the levels and not in the order of their appearance.

```{r}
factor_clarity %>% 
  fct_unique()
factor_clarity %>% 
  unique()
```

## Combine factors
### Combine factors with different levels
With the function `fct_c()` factors with different levels can be combined to one factor covering all levels. First we create two factors. The first one shows the amount of contested disciplines by several countries during the Summer Olympic Games 1896. The second one shows the amount of contested disciplines by countries during the Winter Olympic Games 1924.
```{r}
olympic_1896 <- olympic %>% 
  filter(game == "1896 Summer")

# Create a factor using the default for the `levels` argument in `factor()`.
countries_in_1896 <- factor(olympic_1896$country)

olympic_1924 <- olympic %>% 
  filter(game == "1924 Winter")

# Create a factor using the default for the `levels` argument in `factor()`.
countries_in_1924 <- factor(olympic_1924$country)
```
Note that the default option for the `levels` argument in `factor()` is a sorted set of the given values. As R sorts characters lexicographically, the corresponding levels are the countries sorted alphabetically *from A to Z*.
The factors differ both in their stored values and in their levels. A plot makes this clear:

```{r}
countries_in_1896 %>%
  fct_count() %>%
  ggplot(aes(x = `amount of disciplines`, y = country))+
  geom_col(aes(x = n, y = f))

countries_in_1924 %>%
  fct_count() %>%
  ggplot(aes(x = `amount of disciplines`, y = country))+
  geom_col(aes(x = n, y = f))
```

Now we combine the factors with `fct_c()` and plot it:

```{r}
fct_c(countries_in_1896, countries_in_1924) %>%
  fct_count() %>%
  ggplot(aes(x = `amount of disciplines`, y = country))+
  geom_col(aes(x = n, y = f))
```

Both the underlying values and the levels were combined into one factor.

### Standardise levels of various factors
With the function `fct_unify()` the levels of different factors can be standardised. The resulting output factors will retain their values, but all will have the same levels. Please note, both input and output are in list form.
```{r}
factor_list <- fct_unify(list(countries_in_1896, countries_in_1924))

# Plot the first output factor.
factor_list[[1]] %>%
  fct_count() %>%
  ggplot(aes(x = `amount of disciplines`, y = country)) + 
  geom_col(aes(x = n, y = f))

# Plot the second output factor.
factor_list[[2]] %>%
  fct_count() %>%
  ggplot(aes(x = `amount of disciplines`, y = country)) + 
  geom_col(aes(x = n, y = f))
```

In this case, the underlying values were left unchanged for both factors, but the levels were standardised. This is especially useful when comparing the values of two different factors. 

## Order of levels
### <a name="manual-reordering"></a>Manual reordering of levels
With the function `fct_relevel()` the levels of a factor can be manually reordered. In contrast to the function `levels()`, which only allows the renaming of factor levels, `fct_relevel()` also adjusts the order of the levels themselves, that is the way they are stored internally. An example should clarify this.

```{r}
# Have a look at the original levels.
factor_clarity %>% 
  fct_count()

# Reorder levels according to `new_oder`.
new_order <- c("c1","c3","c5","c7","c2","c4","c6","c8")

factor_clarity %>%
  fct_relevel(new_order) %>% 
  fct_count()
```

Keep in mind, only the function `fct_relevel()` allows a correct manual releveling, while `levels()` only allows to change the levels' names.

### Reordering by frequency
Especially for plots it is often useful to orient the order of the levels on the frequency of the corresponding values. The function `fct_infreq()` allows exactly this. Plotting the unsorted factor leads to a difficult readability.

```{r}
countries_in_1896 %>%
  fct_count() %>%
  ggplot(aes(x = `amount of disciplines`, y = country)) + 
  geom_col(aes(x = n, y = f))
```

The better approach is to sort the data before plotting.

```{r}
# Insert `fct_infreq()` to get the data ordered.
countries_in_1896 %>%
  fct_infreq() %>%    
  fct_count() %>%
  ggplot(aes(x = `amount of disciplines`, y = country)) + 
  geom_col(aes(x = n, y = f))
```

### Reordering by appearance 
The package forcats offers with the function `fct_inorder()` the possibility to orientate the order of the levels of a factor to the occurence of the corresponding values in the data set. To make this clear, we take a look at the first 15 entries in the corresponding data set:

```{r}
olympic_1896$country %>%
  head(15)
```

Create a factor with levels in the order of the appearance shown above:

```{r}
olympic_1896$country %>%
  fct_inorder() %>%
  levels()
```
In this example the same result can be obtained using the function `as_factor()`. However the function `fct_inorder()` can also be applied to already existing factors.

### Reverse level order
The function `fct_rev()` reverses the existing order of the levels of a factor. First have a look at the original order:
```{r}
# Insert `fct_infreq()` to get the data ordered.
countries_in_1896 %>%
  fct_infreq() %>%        
  fct_count() %>%
  ggplot(aes(x = `amount of disciplines`, y = country)) + 
  geom_col(aes(x = n, y = f))
```

Now we reverse the order:

```{r}
# Insert `fct_infreq()` to get the data ordered.
countries_in_1896 %>%
  fct_infreq() %>%
# Insert `fct_rev()` to reverse this order.
  fct_rev() %>%           
  fct_count() %>%
  ggplot(aes(x = `amount of disciplines`, y = country)) + 
  geom_col(aes(x = n, y = f))
```

### Shift levels
The argument `n` in the function `fct_shift()` allows to shift the levels to the left (right) for negative (positive) integer values of `n`, wrapping around end.
Thus a value of `n = 1L` (`n = -1L`) would shift the order of the levels to the right (left) by one location.

```{r}
# Insert `fct_infreq()` to get the data ordered.
countries_in_1896 %>%
  fct_infreq() %>%
# Insert `fct_shift()` to shift levels.
  fct_shift(n = -1L) %>%  
  fct_count() %>%
  ggplot(aes(x = `amount of disciplines`, y = country)) + 
  geom_col(aes(x = n, y = f))
```

### Randomly shuffle levels
The level of a factor can also be randomly shuffled using `fct_shuffle()`. The input argument can be either a factor or a character vector, whereas the output will be a factor. By way of example, this is demonstrated using the factor `countries_in_1896`:
```{r}
# The original levels are sorted alphabetically.
countries_in_1896 %>%
  levels() 

# The shuffled levels are randomly sorted.
countries_in_1896 %>%
  fct_shuffle() %>% 
  levels()  
```
### Reordering levels by other variables
The functions presented in this section bare great similarity to the `fct_relevel()` function introduced in the beginning of this section.
`fct_relevel()` allows for a direct manipulation of the levels by passing the new order to the `levels` keyword argument. In this regard `fct_reorder()` and `fct_reorder2()` are different. Here the levels are reordered according to the result of a function applied to one vector `x` in case of `fct_reorder()` and two vectors `x` and `y` in case of `fct_reorder2()`.

**Note:** We want to point out a few things that have to be kept in mind when using these functions

: 1. The length of the factor `f` to be reordered has the be equal to the lenght of the vector(s) passed to the function. 
That is an error will be provoked, if `length(f) != length(x)` or `length(f) != length(y)`.
2. In case the function returns the same value for two different elements of the vector(s), R will arrange the levels automatically.
3. In case there are multiple occurences of an element, the level assigned to this element will be based on the result of the function for the first corresponding elements in `x` (and `y`).

The example below is intended to unravel this behavior for `fct_reorder()`. We count the occurences of countries in the `olympic_1896` dataset and create a factor where the levels are sorted based on the number of occurences.
```{r}
disciplines_per_country_in_1896 <- olympic_1896 %>% 
  count(country)

# Reorder the countries according to the number of disciplines.
disciplines_per_country_in_1896$country %>%
  fct_reorder(disciplines_per_country_in_1896$n, .desc = TRUE)
```

## Change the value of levels
### Renaming the levels 
Let's say you want to change the name of the levels (which also impies changing the corresponding value elements) because you are unhappy with the current naming. `fct_recode()` allows to manually assign new names to certain levels, without affecting the order of the levels or having an impact on levels that are not included in the function call. 
The `diamonds$clarity` column is an ideal example for the use of `fct_recode()`. For you conviniece we repeat some of the steps made in the beginning of this chapter:
```{r}
# Define the correct order of the levels in ascending order.
levels_clarity <- c("I1", "SI2", "SI1", "VS2", "VS1", "VVS2", "VVS1", "IF")
# Make `diamonds$clarity` a factor and assign the correctly ordered levels.
factor_clarity <- factor(x = diamonds$clarity, levels = levels_clarity)

# Since the level notation is somewhat cryptic, we want to change it.
renamed_factor_clarity <- factor_clarity %>% 
  fct_recode(`Included 1 (worst)`="I1",
             `Slightly Included 2`="SI2",
             `Slightly Included 1`="SI1",
             `Very Slightly Included 2`="VS2",
             `Very Slightly Included 1`="VS1",
             `Very Very Slightly Included 2`="VVS2",
             `Very Very Slightly Included 1`="VVS1",
             `Internally Flawless (best)`="IF")

# Show the renamed levels.
renamed_factor_clarity %>% 
  levels()

```
Note, that this approach is in fact similar to the steps described in  [5.2.3  Inspect and set levels](#Inspect-and-set-levels).

<a name="recode-2"></a>One could also combine multiple levels into one using `fct_recode()` as shown in the example below, where the numerical distinction within the GIA categories is dropped, that is *Slightly Included 2* and *Slightly Included 1* will be combined into the level *Slightly Included*, etc.
```{r}
# Since the level notation is somewhat cryptic, we want to change it
# and drop the numerical distinction within each category.
renamed_factor_clarity_2 <- factor_clarity %>% 
  fct_recode(`Included (worst)`="I1",
             `Slightly Included`="SI2",
             `Slightly Included`="SI1",
             `Very Slightly Included`="VS2",
             `Very Slightly Included`="VS1",
             `Very Very Slightly Included`="VVS2",
             `Very Very Slightly Included`="VVS1",
             `Internally Flawless (best)`="IF")

# Show the combined levels.
renamed_factor_clarity_2 %>% 
  levels()

```
The resulting `renamed_factor_clarity_2` now has only five distinct levels as compared to eight distinct levels in `renamed_factor_clarity`.

Another scenario might be that we want to add either a pre- or a suffix or both to the current level or change the levels in some other general manner. To this end, the forcats package provides the `fct_relabel()` function. Its second argument (or its first argument in case of piping) is a function that has to return a character vector. The `countries_in_1896` factor variable is used to showcase how `fct_relabel()` works.
```{r}
add_prefix <- function(input_char) {
  paste("Country:", input_char)
}
# The `add_prefix()` function is applied to each level 
# returning a character vector of the new labels. 
prefixed_countries_in_1896 <- countries_in_1896 %>%
  fct_relabel(add_prefix)

# Show the combined levels.
prefixed_countries_in_1896 %>% 
  levels()
```

### Anonymize levels
There might be situations, especially in the context of data privacy, where you want or have to anonymize your factor data. Assigning numeric IDs is a good way to do this and `fct_anon()` does exactly this. It randomly assigns integer values (converted to characters) starting from `1` to the levels of a factor variable. The largest integer value to be assigned depends on the number of distict levels in the variable. In the following example the `countries_in_1896` factor variable with 12 distinct levels is randomly anonymized.
```{r}
# The function `fct_anon()` additionally allows to
# define a prefix for the new random integer levels.
countries_in_1896 %>% 
  fct_anon(prefix="id") %>% 
  head(25)
```

### Collapse multiple levels into one
The `fct_collapse()` function provides essentially the same functionality as displayed in the [2nd example for `fct_recode()`](#recode-2).
The syntax however is slightly different as the levels to be combined are specified in a single vector as shown in the example below.
```{r}
renamed_factor_clarity_3 <- factor_clarity %>% 
  fct_collapse(`Included (worst)`="I1",
               `Slightly Included`= c("SI2", "SI1"),
               `Very Slightly Included`= c("VS2","VS1"),
               `Very Very Slightly Included`= c("VVS2", "VVS1"),
               `Internally Flawless (best)`="IF")
```
It should be noted, that `fct_recode()` could be easily replaced with `fct_collapse()` in the [2nd example for `fct_recode()`](#recode-2) as `fct_collapse()` also works with single element character vectors. Replacing `fct_collapse()` with `fct_recode()` in the example above on the other hand is not possible, because `fct_recode()` cannot work with multiple element vectors such as `c("SI2", "SI1")`. In this regard, we rather recommend using `fct_collapse()` than `fct_recode()` for combining different levels.

### Aggregate levels into a lump 
This topic is related to the rationale behind `fct_collapse()` described in the previous section. In contrast to specifying the levels to be combined explicitly, the forcats package also offers the possibility to lump levels together. That is combining several levels together to a single lump level called `Other` by default  There are several different functions that allow lumping based on different criteria.  
The functions which is probably easiest to grasp is `fct_lump_n()` which takes an integer value `n` as the second (or the first argument in case of piping). 
It allows to preserve the `n` most (least) frequent levels for `n>0` (`n<0`). All other levels will be lumped together into one level that has the lowest order and referred to as `Other` by default.
Once again, we use data from the Summer Olympic Games in 1896 to give an example.
```{r}
# Only keep the three most frequent countries 
# and lump all other countries into `Other`.
countries_in_1896 %>% 
  fct_lump_n(3) %>% 
  levels()
```
We just discovered that Germany was one of the countries competing in the most disciplines during the Summer Olympic Games 1896 albeit quite laboriously.
Note that we can however not conclude that most disciplines were competed by German athletes in 1896 as the levels are ordered alphabetically (except for `Other` which will always be last) and not by the number of disciplines per country. In fact, with 148 disciplines competed by Greek athletes, 94 by German and 27 by athletes from the US, Germany ranks second.

Let's go one step further and assume we are only interested in the countries whose athletes competed in at least 10 disciplines in the Summer Olympic Games 1896.
For that purpose we call the `fct_lump_min()` function specifying the threshold of 10. 
```{r}
# Only keep the countries whose athletes competed in at least 10 disciplines 
# and lump all other countries.
countries_in_1896 %>% 
  fct_lump_min(10) %>% 
  levels()

```
Now we know that there were only eight countries whose athletes competed in at least 10 disciplines.

We could also be interested in relative rather than absolute number, say countries that provide at least a 10\% share of the total number of competed disciplines which corresponds to 38 competed disciplines in 1896 (see above). To figure this out, we lump proportionally using `fcr_lump_prop()`.
```{r}
# Only keep the countries whose athletes are responsible for 
# at least 10% of all competition records
# and lump all other countries.
countries_in_1896 %>% 
  fct_lump_prop(0.1) %>% 
  levels()
```


### Manually lump levels
In the previous section, we have combined levels into a lump called `Other` based on numerical criteria. In addition to this, we can also create our own lump using `fct_other()`.In this function, we specify the levels we want to keep (drop) as a vector to the `keep` (`drop`) keyword argument. 
The following code demonstrates how to only keep the levels `Denmark` and `Australia` in the `countries_in_1896` factor variable.
```{r}
# Only keep the levels `Denmark` and `Australia`.
countries_in_1896 %>% 
  fct_other(keep=c("Denmark", "Australia")) %>% 
  levels()

# Keep all levels except for `Denmark` and `Australia`.
countries_in_1896 %>% 
  fct_other(drop=c("Denmark", "Australia")) %>% 
  levels()
```
Note that the latter example is equivalent to using `fct_collapse(Other=c("Denmark", "Australia"))`.


## Add or drop levels
### Add levels
In order to add levels to a factor variable we can use the `fct_expand()` function, which takes the level as an argument. Let's switch to the `diamonds` dataset one more time as we want to use all GIA grades as levels now.
```{r}
# The additional GIA grades which do not exist in the diamonds dataset.
additional_GIA_grades <- c("FL", "I2", "I3")

# Expand the original factor by the additional levels.
expanded_factor_clarity <- factor_clarity %>%
  fct_expand(additional_GIA_grades)

# Note that the new levels are appended.
expanded_factor_clarity %>% 
  levels()
```
Note that the additional levels get appended which is not necessarily what we want. Thus an additional manual reordering of the levels as describe in [5.4.1 Manual reordering of levels](#manual-reordering) is required here:
```{r}
# All GIA grades in the correct order.
all_GIA_grades <- c("FL", "IF", "VVS1", "VVS2", "VS1", "VS2", "SI1", "SI2", "I1", "I2", "I3")

# Reorder the levels of the expanded factor.
reordered_expanded_factor_clarity <- expanded_factor_clarity %>%
  fct_relevel(all_GIA_grades) 

reordered_expanded_factor_clarity %>% 
  levels()
```
A less tedious approach would be to include the additional levels already in the definition of the factor variable using `factor()` as shown in the following.
```{r}
factor(x = diamonds$clarity, levels=all_GIA_grades) %>%
  levels()
```
### Drop levels
Unfortunately, we haven't found any flawless diamond, so the level `FL` isn't used. To drop unused levels, the forcats package provides a funtion called `fct_drop()`.
```{r}
# Only drop the unused "FL" level.
reordered_expanded_factor_clarity %>% 
  fct_drop("FL") %>% 
  levels()

# Drop all unused levels when no argument is used.
reordered_expanded_factor_clarity %>% 
  fct_drop() %>% 
  levels()
```

### Assign a level to `NAs`
Usually when a vector contains `NAs` they are omitted in the conversion to a factor variable. The forcats package allows to explicitly assign a level to these values using the `fct_explicit_na()` function. To give an example, we return to the `olympic_1896` dataset but we focus on the `city` column this time.
```{r}
# Convert the `medal` column into a factor.
olympic_1896$medal %>% 
  factor() %>% 
  fct_count()

# Convert the `medal` column into a factor, replacing the NA values.
olympic_1896$medal %>% 
  fct_explicit_na("No medal") %>% 
  fct_count()
```


