# Dates and times: lubridate {#lubridate}

Resources:

- [Lubridate homepage](https://lubridate.tidyverse.org/)
- [Cheatsheet](https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf)
- [Book Chapter in R4DS](https://r4ds.had.co.nz/dates-and-times.html)
- [Vignette](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html)

Suggested data set: weather-kiel-holtenau

## What is Lubridate?
Lubridate is an R-Package designed to ease working with date/time variables. These can be quite challenging in baseR and lubridate allows for frictionless working with dates and times, hence the name.

Lubridate ist part of the tidyverse package, but can be installed seperately as well. It probably reveals most of its usefullness in collaboration with other tidyverse packages. A useful extension, depending on your data, might be the time-series package which is not part of tidyverse.

All mentioned packages can be optained with the following commands. 

TODO @Robert funktioniert bei mir nicht
package.install("lubridate")
package.install("tidyverse")

TODO @Robert in use?
package.install("time-series")

If the have been installed previously in your environment, they might have to be called upon by using 
library(tidyverse) and so forth - see code chunks below.

## Basics

Some examples of real world date - time formats found in datasets:

How people talk about dates and times often differs from the notation of the given information. Depending on the specific use of the data, the given information might be more or less granular. When people in the USA talk distance between two places, they often give an approximation of how long it will take a person to drive from A to B and round-up or down to the hour. 

Flight schedules will most likely be exact to the minute, while some sensordata will probably need to be exact to the milisecond. So there will be differing granularity in date and time data. 

Even if this would not be a challenge, we still would have to deal with different notations of date and time. People in Germany will write a day-date like: dd.mm.yyyy or short dd.mm.yy, while the anglo-saxon realm will use mm.dd.yyyy frequently and the most chronologically sound way would be to use yyyy.mm.dd, but this doesn't seem to stick with humans. 

On top of these issues there's the fact that time itself does not make the impression of being the most exact parameter out there. Universal time might appear linear, but the way our planet revolves our galaxy has made it neccessary to adjust date and times every now and then, so our Kalender stays in tune with our defined seasons. This creates leap years, skipped seconds, daylight-savings time and last, but not least time-zones, which can mess things up even further.

TODO @Robert - Wollen die noch mehr beschreiben?
Three types of date/time data
Sys.time() functions
Time Formats

## Import data, clean date-time
We will apply the lubridate package to Weather data from on stationary sensor in northern Germany, the weather station in Kiel-Holtenau to be more exact.

Before we introduce the library, the prerequisites must be created.
```{r libararies, echo=FALSE}
library(readr)          # part of the tidyverse
library(lubridate)      # the mentioned lubrication for date-time wrangling
library(tidyverse)      # the tidyverse with its dplyr functions for data wrangling
library(ggplot2)        # data visualisation package (is actually part of tidyverse, but still)
```
A first step is to import the data, which is given in a csv-format. We will use the tidyverse version of read_csv to accomplish this step.

The data will be called df (data.frame) in order to make reference in code an writing more efficient further down.

This is the simple approach to read a file with suffix csv.
```{r read_csv simple, echo=FALSE}
df <- read_csv("data/weather_kiel_holtenau.csv")
head(df,5)
```
All columns are recognized as double values.

By looking at the "MESS_DATUM" column it becomes apparent that this represents a timestamp that was created every 10 minutes. The standard column type definition of the import tool have´nt sufficed to format this appropriately, which is why the column will be defined as a double for now.

Using the code generator of the data import tool it is possible to format the variables in the appropriate classes. The code snippet is automatically generated by the import tool. For our concern of the timestamp variable please note the part where the column MESS_DATUM = col_datetime(format = "%Y%m%d%H%M") which will reach our present goal to format the timestamp and class as a POSIXct.

```{r read_csv ideal, echo=FALSE}
df <- read_csv("data/weather_kiel_holtenau.csv", 
          col_types = cols(MESS_DATUM = col_datetime(format = "%Y%m%d%H%M"), 
                                                   NIEDERSCHLAGSDAUER = col_integer(), 
                                                   NIEDERSCHLAGSINDIKATOR = col_integer(), 
                                                   STATIONS_ID = col_integer()))
head(df,5)
```
We will examine the data further down below. At this point it should be mentioned again that we generated the code to import the data by using the import readr tool of RStudio. This allows a first look at the raw csv data before import and some tweaking of the variables and their classes. We left the locale portion of the import tool untouched.

The following lines of Code import the same data in the MESS_DATUM column as a string of characters again and then calls a specific lubridate function (ymd = year-month-day) which basically recognizes the time format in the column given only very little information. In our case we only specify the order in which the date-time information is given, which is Year-Month-Day-Hour-Minute. The ymd_hm function delivers the desired result however, only if it is applied to a character string, which is why we chose to overwrite the dataset df with a new import procedure that makes sure that the column MESS_DATUM is given as a character after import.
```{r read_csv alternate 1, echo=FALSE}
df1 <- read_csv("data/weather_kiel_holtenau.csv", 
          col_types = cols(
            MESS_DATUM = col_character(), 
            NIEDERSCHLAGSINDIKATOR = col_integer(), 
            STATIONS_ID = col_integer()))
df1$MESS_DATUM <- ymd_hm(df1$MESS_DATUM)
head(df1,5)
```
As can be seen above, the MESS_DATUM is now in a POSIXct format again, which is what will be needed for further calculations and analysis of the data set.

Alternatively you could use the following parsing function to achieve the same result. Note that the parse_date_time function also needs the timestamp to be in character format.
```{r read_csv alternate 2, echo=FALSE}
df2 <- read_csv("data/weather_kiel_holtenau.csv", 
    col_types = cols(MESS_DATUM = col_character()))
df2$MESS_DATUM <- parse_date_time(df2$MESS_DATUM, orders ="Ymd HM")
head(df2,5)
```
OK we have successfully formated the time-stamp data into a productive date-time (dttm) format.

## Check and modify data
The next steps are a check and elimination procedure to eliminate missing values (NAs) from the dataset,
if some observations might have failed to generate data successfully.
```{r is NA in MESS_DATUM, echo=FALSE}
df %>% 
  filter(is.na(MESS_DATUM)) %>%                   # Checking if there are observations with a missing MESS_DATUM
  head()
```
We don't encounter any NAs in the MESS_DATUM column. Eliminating NAs from the MESS_DATUM Column including the other variables of that specific observations from the data frame would have been possible with: 
```{r is no NA in MESS_DATUM, echo=FALSE}
df <- df %>% 
  filter(!is.na(MESS_DATUM))                      # missing MESS_DATUM observations (NAs) would/will be excluded from the data frame
```
OK. Time to have first explorative look at the data set.
```{r max/min/range MESS_DATUM, echo=FALSE}
max(df$MESS_DATUM)                                # The latest observation is dated 10 minutes before midnight on the 13th of April 2020
min(df$MESS_DATUM)                                # The earliest observaation is dated 14th of April 2019
range(df$MESS_DATUM)                              # Another way to get the same information
```
To gather some information on the other variables we can use the range function again.
```{r range of other columns, echo=FALSE}
range(df$STATIONS_ID)                             # This confirms: We are only dealing with data from one and only one weather sensor in the whole data set.
range(df$TEMPERATUR)                              # We have a range of temperatues between -2.5 and +32.1 degrees  celius, which sounds about plausible.
range(df$RELATIVE_FEUCHTE)                        # Delivers an unexpected range, where the minimum is -999 and the maximum is +100. Both values are either not possible (-999) or only of theoretic value (+100)
range(df$NIEDERSCHLAGSDAUER)                      # This delivers an implausible minimum of -999 and an expected maximum of 10 (the maximum amount of minutes in a 10 minute interval).
range(df$NIEDERSCHLAGSHOEHE)                      # Again the range shows a minimum at -999 and a maximum of 7.85 which has to be interpreted as mm which equals litres per square meter (6 x 7.85 = 47.1/h = torrential)
range(df$NIEDERSCHLAGSINDIKATOR)                  # A look at the data reveals a binary 1 or a 0 for this variable, i.e. it rained or it didn't. -999 must be interpreted as a failed observation
```
We found observations in a number of variables that would need to be deleted from the table to get operable data.
Theoretically, the following columns are only relevant if the NIEDERSCHLAGSINDIKATOR is set to 1 = rainfall.  This column can be considered like a switch for the other two rain-connected columns:
- NIEDERSCHLAGSDAUER
- NIEDERSCHLAGSHOEHE

Getting an idea about the extent of failed observations can be accomplished with the following code snippet:
```{r correct NIEDERSCHLAGSINDIKATOR - before, echo=FALSE}
values_of_ind <- df %>% 
  group_by(NIEDERSCHLAGSINDIKATOR) %>%
  tally()
head(values_of_ind)
```
So 127 observations of rainfall are faulty. To modify the NIEDERSCHLAGSINDIKATOR for upcoming analyzes we do the following in order to set the -999 Values to a more neutral Zero:
```{r correct NIEDERSCHLAGSINDIKATOR - action, echo=FALSE}
df$NIEDERSCHLAGSINDIKATOR <- ifelse(df$NIEDERSCHLAGSINDIKATOR == -999, 0, df$NIEDERSCHLAGSINDIKATOR)
value_of_ind_n <- df %>% 
  group_by(NIEDERSCHLAGSINDIKATOR) %>%
  tally()
head(value_of_ind_n)
```
The column RELATIVE_FEUCHTE must also be adjusted in the same way.
```{r correct RELATIVE_FEUCHTE - before, echo=FALSE}
values_of_rel <- df %>% 
  group_by(RELATIVE_FEUCHTE) %>%
  filter(RELATIVE_FEUCHTE < 0) %>%
  tally()
head(values_of_rel)
```
The negative values are set to 0 again.
```{r correct RELATIVE_FEUCHTE - action, echo=FALSE}
df$RELATIVE_FEUCHTE <- ifelse(df$RELATIVE_FEUCHTE == -999, 0, df$RELATIVE_FEUCHTE)
values_of_rel_n <- df %>% 
  group_by(RELATIVE_FEUCHTE) %>%
  tally()
head(values_of_rel_n)
```
The column NIEDERSCHLAGSHOEHE can also be tested and adjusted in the same way.
It is not necessary because the negative values can be hidden in the further analyzes using the NIEDERSCHLAGSINDIKATOR = 1.
```{r}
values_of_ARN <- df %>% 
  group_by(NIEDERSCHLAGSHOEHE) %>%
  filter(NIEDERSCHLAGSHOEHE < 0) %>%
  tally()
head(values_of_ARN)
```
```{r}
df$NIEDERSCHLAGSHOEHE <- ifelse(df$NIEDERSCHLAGSHOEHE == -999, 0, df$NIEDERSCHLAGSHOEHE)
values_of_RFH <- df %>% 
  group_by(NIEDERSCHLAGSHOEHE) %>%
  tally()
head(values_of_RFH)
```
We have set Variable NIEDERSCHLAGSHOEHE to 0 in these 129 oberservations. In a total of 47,753 observations with 0 as a value the distortion this will cause can be called negligible.

## Before exploration
We now have clean and tidy data and can begin with exploration and interpretation.

The "TEMPERATUR" is given in degrees Celsius, the "RELATIVE-FEUCHTE" is a percentage Value for humidity which refers to the degree of water saturation that is prevalent in the air at a given temperature. As temperature increases, the air can absorb larger amounts of water, hence the relativity of this variable.

The next variable is "NIEDERSCHLAGSDAUER" which is given as an integer smaller or equal to 10. Therefore it gives the time it has rained during the timestamp intervall of 10 Minutes.
The variable "NIEDERSCHLAGSHOEHE" is a measure of rainfall intensity. Its maximum value can also be checked by the following command:
TODO @Marcus / @Robert - Entfernen Auswertung jetzt mit Indikator
```{r max NIEDERSCHLAGSHOEHE NA, echo=FALSE}
max(df$NIEDERSCHLAGSHOEHE, na.rm = FALSE)
```
We can assume that this number gives us the amount of rainfall in milimeters, which is a common definiton and is equivalent as liters of rainfall per squaremeter in a given time intervall. A strong rainfall in central Europe can be expected to generate around 30mm/h of rainfall, i.e. 30 liters of rainfall per squaremeter per hour. The maximum value is therefore an indicator of a very heavy downpour as a continuation for a whole hour would have yielded 6 x 7.85 = 47.1 litres per hour.

The next variable is simply a binary expression of rainfall (1) or no rainfall (0) in the given intervall. This is relevant to measure as some types of rainfall do not generate enough water to messure an amount of water. The dreaded Northgerman "drizzle" comes to mind.

Again: Since the sensor has taken a snapshot every 10 minutes, we have six observations per hour.

### Intervals
Lubridate allows for the definition of an object with the "interval" class. The intervall is simply defined as the time elapsed between two points on the date-time line.
The interval of a single day - from the first of March to the second of March in the following example can be defined as follows:
```{r interval TAG, echo=FALSE}
tag_int <- interval(ymd("2020-03-01"), ymd("2020-03-02"))
class(tag_int)
```
A small interval of one day turned out to be good for the further analysis.
This makes it easier to check the results at a glance.

Only when the expected results have been achieved for a short period do we consider the next largest period, e.g. Week, month, quarter and year.

TODO @MARCUS - Dein folgender Code würde zumindest schnell eine durchschnittsstemperatur für die jeweilige Jahreszeit liefern. Es müsste dann halt noch in einer Tabelle zusammengeführt werden.
Hier könnte jetzt die Betrachtung von Intervallen als Anwendung auf die Jahreszeiten folgen.

### PERIODS and DURATIONS
For any weather data, an analysis fo seasonal differences is a natural (excuse the pun) objective.
The beginning of each years seasons align with the solar incidences, i.e. the spring and autumnal equinoxes an the days of most and least sunlight hours in summer and winter.
The following code snippets define the starting points of the different seasons as points on the POSIXct timeline and use these points to calculate new points in time by adding durations to them.

For a seasons view, we create the appropriate times.
```{r create durations}
season_19_spring <- ymd_hm("2019-03-20 22:58", tz="CET")
season_19_summer <- ymd_hm("2019-06-21 17:54", tz="CET")
season_19_autumn <- ymd_hm("2019-09-23 09:50", tz="CET")
season_19_winter <- ymd_hm("2019-12-22 05:19", tz="CET")
season_20_spring <- ymd_hm("2020-03-20 04:49", tz="CET")
season_20_summer <- ymd_hm("2020-06-20 23:43", tz="CET")
```
The information of seasons are [from ](https://www.timeanddate.de/astronomie/jahreszeiten "timeanddate.de")

Our target is to create an interval spring.
The following information are available to determine the end point of spring:
**Frühling** 2020
Beginn (Tag-und-Nachtgleiche März)
20. März 04:49
Dauer 92 Tage, 17 Std, 54 Min
**Sommer** 2020
Beginn (Sonnenwende Juni)
20. Juni 23:43
Dauer 93 Tage, 15 Std, 46 Min
```{r}
print(paste("spring 2020 - start: ",season_20_spring))
print(ddays(92))                             # create a DURATIONS
class(ddays(92))
typeof(ddays(92))
season_20_spring_dur <- season_20_spring + 
                          ddays(92) + 
                          dhours(17) + 
                          dminutes(54)
print(season_20_spring_dur)
class(season_20_spring_dur)
typeof(season_20_spring_dur)
print(paste("spring 2020 - end with DURATIONS:",season_20_spring_dur, "expected:", season_20_summer, "is failure:", season_20_spring_dur!=season_20_summer))

print(days(92))                              # create a PERIODS
class(days(92))
typeof(days(92))
season_20_spring_p <- season_20_spring + 
                        days(92) + 
                        hours(17) + 
                        minutes(54)
print(season_20_spring_p)
class(season_20_spring_p)
typeof(season_20_spring_p)
print(paste("spring 2020 - end with PERIODS:",season_20_spring_p, "expected:", season_20_summer , "is failure:", season_20_spring_p!=season_20_summer))
```
The example shows that the use of PERIODS or DURATIONS must be weighed up and balanced for the challenge you are looking for. 
The periods-example has a difference of one hour to the durations-example.

The seasons summer, autumn and winter are ideal for our data frame.
```{r}
int_season_summer <- interval(season_19_summer, season_19_autumn - minutes(1))
int_season_autumn <- interval(season_19_autumn, season_19_winter - minutes(1))
int_season_winter <- season_19_winter %--% (season_20_spring - minutes(1))
```

### Groups
The following GROUP_BY command creates a unique grouping per hour and then calculates the average temperature for every hour, then pastes this value into each observation per hour.
```{r create df_group of TAG for STUNDE/TAG - simple, echo=FALSE}
df_sel <- df %>%
  filter(MESS_DATUM >= int_start(tag_int) & MESS_DATUM <= int_end(tag_int))
df_group <- df_sel %>% 
  group_by(STATIONS_ID , hour(MESS_DATUM) , day(MESS_DATUM) ) 
head(df_group, 10)
```
For better readability, the columns can be renamed after a Group-By with rename function.
```{r create df_group of TAG for STUNDE/TAG - rename, echo=FALSE}
df_group <- df_sel %>% 
  group_by(STATIONS_ID , hour(MESS_DATUM) , day(MESS_DATUM) ) %>% 
  rename( "STUNDE" = 'hour(MESS_DATUM)', "TAG" = 'day(MESS_DATUM)' ) 
head(df_group, 10)

```
Or the columns can be named directly in the Group-By command:
```{r create df_group of TAG for STUNDE/TAG - rename - ideal, echo=FALSE}
df_group <- df_sel %>% 
  group_by(STATIONS_ID , STUNDE = hour(MESS_DATUM) , TAG = day(MESS_DATUM) ) 
head(df_group, 5)

```
Now we can filter the data by year, month, week, day and hour of the day. This should give us possibilities to aggregate freely over any time interval.

#### Seasons
Now we create data frames for analyze seasons.
```{r select data of summer, echo=FALSE}
df_sea_sel_summer <- 
  df %>%
    filter(MESS_DATUM >= int_start(int_season_summer) & MESS_DATUM <= int_end(int_season_summer)) 
head(df_sea_sel_summer)

df_sea_sel_autmn <- 
  df %>%
    filter(MESS_DATUM >= int_start(int_season_autumn) & MESS_DATUM <= int_end(int_season_autumn)) 
df_sea_sel_winter <- 
  df %>%
    filter(MESS_DATUM >= int_start(int_season_winter) & MESS_DATUM <= int_end(int_season_winter)) 
#df_group <- df_sel %>% group_by(STATIONS_ID , hour(MESS_DATUM) , day(MESS_DATUM) ) head(df_group, 10)
```
TODO @Marcus/@Robert mess_datum bei Sommer ab 18:00, wenn 17:54

## Exploration - Analysis

First we are going to calculate average temperatures for different standard intervalls like hours, days, weeks, months. We will visualise some data and eventually calculate new variables or aggregates for humdity, rainfall etc. as well, plus more insights, that are might not be apparent at this stage.

### Temperatur
The first variable of interest should be "TEMPERATUR". A quick visualisation delivers this picture:
```{r ggplot MESS_DATUM TEMPERATUR for JAHR, echo=FALSE}
ggplot(df)+
  geom_point(aes(MESS_DATUM, TEMPERATUR), colour = "red", size = 0.1)
```
This is a representation of all 52,704 observations of temperature and naturally appears quite crowded. 
However, a typical course of the seasons during a year can already be interpreted from this plot.

ACHTUNG - ENTWEDER IN BESSER ODER RAUS! IST NICE TO HAVE!
Versuch des Plottings einer Regression. Hier entsteht natürlich eine Regressionslinie durch alle Punkte, was keinen Sinn ergibt. 
Es müsste als eine Regression auf eine Zeitreihe definiert bzw. vermutlich auf rollierende Durchschnitte berechnet werden.
Mit einem vernünftigen rollierenden Durchschnitt kann so eine regression inkl. Standard-Error aber auch eine schöne Visualiserung abgeben (ggf. sogar ohne die Datenpunkte selbst)

```{r ggplot MESS_DATUM TEMPERATUR, echo=FALSE}
ggplot(df, aes(MESS_DATUM, TEMPERATUR))+
  geom_smooth(method = "lm", se=TRUE, color="black", formula = y ~ x) +
            geom_point(color = "red", size = 0.1)
```
Let'S try to get a clearer picture of the temperature during the course of the observed year. 
We need to form averages and aggregates to make visualisation more to the point and get less crowded pictures.

New variables should represent other dimensions of date-time data. So we could produce variables for hours, 24hdays, weeks, months and even possibly seasons.
Luckily lubridate has functions for us that will generate these new variables:

Assigning a year to every observation by creating a new column with lubridate function "year":
```{r mutate MESS_DATUM - JAHR, echo=FALSE}
df <- df %>% 
  mutate(JAHR = year(df$MESS_DATUM))
```
Assigning a month to every observation by creating a new column with lubridate function "month"
```{r mutate MESS_DATUM - MONAT, echo=FALSE}
df <- df %>% 
  mutate(MONAT = month(df$MESS_DATUM))
```
Assigning an Epiweeknr (EKW = Epi Kalender Woche) to every observation by creating a new column with lubridate function "epiweek":
```{r  mutate MESS_DATUM - EKW, echo=FALSE}
df <- df %>% 
  mutate(EKW = epiweek(df$MESS_DATUM))
```
Assigning a daynumber to every observation by creating a new column "JTAG" (Year day):
```{r mutate MESS_DATUM - TAG im Jahr, echo=FALSE}
df <- df %>%
  mutate(JTAG = yday(df$MESS_DATUM))
```
Assigning an hour to every observation by creating a new column "STUNDE"
```{r mutate MESS_DATUM - STUNDE, echo=FALSE}
df <- df %>%
  mutate(STUNDE = hour(df$MESS_DATUM))
```
Our data frame (df) has 12 variables now and thus we can filter the data by year, month, week, day and hour of the day. This gives us possibilities to aggregate.
The following GROUP_BY command creates a unique grouping per hour and then calculates the average temperature for every hour, then pastes this value into each observation per hour.
```{r avg TEMPERATUR STUNDE, echo=FALSE}
df_group_Stunde <- df %>%
  group_by(STATIONS_ID, JAHR, MONAT, EKW, JTAG, STUNDE)

df_av_temp_Stunde  <- df_group_Stunde %>%
  summarise(AVGTEMPH = mean(TEMPERATUR)) %>%
  arrange(STATIONS_ID, JAHR, MONAT, EKW, JTAG, STUNDE)
```
This newly created table has reduced the number of observations by the factor 6 to 8,784 and reveals the average temperature per hour.
Let's plot the data again as hourly temperature averages per day:
```{r ggplot avg TEMPERATUR STUNDE, echo=FALSE}
ggplot(df_av_temp_Stunde)+
  geom_point(aes(JTAG, AVGTEMPH), colour = "red", size = 0.1)
```
We can clearly see that the data has become aggregated and that the temperature picture becomes somewhat more clear.
Also the visualisation now stretches over the course of a whole calender year from left to right.
Let's go on step further and shorten the data down to daily averages by:
```{r avg TEMPERATUR JTAG, echo=FALSE}
df_group_JTAG <- df %>%
  group_by(STATIONS_ID, JAHR, MONAT, EKW, JTAG)

df_av_temp_JTAG  <- df_group_JTAG %>%
  summarise(AVGTEMPD = mean(TEMPERATUR)) %>%
  arrange(STATIONS_ID, JAHR, MONAT, EKW, JTAG) 
head(df_av_temp_JTAG, 10)
```
This newly created table has reduced the number of observations by the factor 24 to 366 (the number of days in a leap year) and reveals the average temperature per day.
Let's plot the data again as dayly temperature averages over the whole period as a line:
```{r ggplot avg TEMPERATUR JTAG, echo=FALSE}
ggplot(df_av_temp_JTAG)+
  geom_line(aes(JTAG, AVGTEMPD), colour = "red", size = 0.1)
```
The data starts to make more visual sense. The key take-aways from this plot, next to the rather trivial finding of higher summer temperatures, 
is the rather high volatility that appears to be attached to daily average temperature throughout the year.

Let's boil the data down to weekly temperature averages:
```{r avg TEMPERATUR WOCHE, echo=FALSE}
df_group_EKW <- df %>%
  group_by(STATIONS_ID, JAHR, MONAT, EKW)

df_av_temp_EKW  <- df_group_EKW %>%
  summarise(AVGTEMPW = mean(TEMPERATUR)) %>%
  arrange(STATIONS_ID, JAHR, MONAT, EKW) 
head(df_av_temp_EKW, 10)
```
and plot again:
```{r ggplot avg TEMPERATUR WOCHE, echo=FALSE}
ggplot(df_av_temp_EKW)+
  geom_line(aes(EKW, AVGTEMPW), colour = "red", size = 0.5)
```
Let's look at monthly averages:
```{r avg TEMPERATUR MONAT, echo=FALSE}
df_group_MONAT <- df %>%
  group_by(STATIONS_ID, JAHR, MONAT)

df_av_temp_MONAT  <- df_group_MONAT %>%
  summarise(AVGTEMPM = mean(TEMPERATUR)) %>%
  arrange(STATIONS_ID, JAHR, MONAT) %>%
  view()
```
and plot again:
```{r ggplot avg TEMPERATUR MONAT, echo=FALSE}
ggplot(df_av_temp_MONAT)+
  geom_line(aes(MONAT, AVGTEMPM), colour = "red", size = 1)
```
Which is the kind of temperature curve we would expect to see in a north German location like Kiel with a clear pattern of 3 months of summer with high average temperatures.
It is interesting to note that April 2020 is apparently much warmer on average than 2019. The dimension indicates a difference of about 2 degrees celsius.
in April 2019 represents about the second half of April and the data collected in 2020 represents just about the first half of April. This means, even though the 2020 data has been 
collected earlier during the course of the year, the average temperature has been significantly higher.

#### Average temperature for a day
In principle, no indicator has to be used to determine the average temperature, since the value is completely in the data frame.
```{r avg TEMPERATUR TAG STUNDE on an interval of a day}
df_avg1 <- df_group %>% 
  summarise(avg = mean(TEMPERATUR)) %>% 
  arrange( STATIONS_ID , TAG , STUNDE) %>%
  view()  
```

### Rainfall
#### Sum with condition
The situation is different with the duration of precipitation. 
The indicator must be used here to determine the duration.

without indicator
```{r sum NIEDERSCHLAGSDAUER - Tag/Stunde on an interval of a day, echo=FALSE}
df_sum1 <- df_group %>% 
  summarise(sum(NIEDERSCHLAGSDAUER)) %>% 
  arrange( STATIONS_ID , TAG, STUNDE) %>%
  view()
```

with indicator and renamed column
```{r sum NIEDERSCHLAGSDAUER - condition - Tag/Stunde on an interval of a day, echo=FALSE}
df_sum2 <- df_group %>% 
  summarise(MENGE = sum(NIEDERSCHLAGSDAUER[NIEDERSCHLAGSINDIKATOR==1])) %>% 
  arrange( STATIONS_ID , TAG, STUNDE) %>%
  view()
```

(Idea - check, if data is openly available as well. This could lead to an excurse about how to generate a raw dataset from this weatherstation.)

### Humidity
What we can say at this point is that the temperatures in the City of Kiel have a clear seasonal pattern, but remain highly volatile intra-day and inter-day.
Weather however is not just defined as temperature, but humidity and precipitation play an integral role of our sensation and definition of weather as well.

Let's look at humidity the same way we looked at temperatures as hourly averages:

```{r avg RELATIVE_FEUCHTE, echo=FALSE}
df_av_humi_Stunde  <- df_group_Stunde %>%
  summarise(AVGHUMI = mean(RELATIVE_FEUCHTE)) %>%
  arrange(STATIONS_ID, JAHR, MONAT, EKW, JTAG, STUNDE)
```
Let's plot the data again as hourly humidity averages per day:
```{r ggplot avg RELATIVE_FEUCHTE, echo=FALSE}
ggplot(df_av_humi_Stunde)+
  geom_point(aes(JTAG, AVGHUMI), colour = "red", size = 0.1)
```
ACHTUNG: in dieser Ansicht zerstören die -999 und offenbar auch noch ein paar -125er Werte die Ansicht. Das müsste sich nach MARCUS Bereinigungsmethode ja erledigen.
Dann auch hier Tages, Wochen, Monats und Jahreszeitendurchschnitte berechnen und zeigen.

Im nächsten Schritt wären dann die Niederschläge zu zeigen. Hier haben wir es dann nicht mehr mit durchschnitten, sondern Summen (machen Stunden Sinn?) an Niederschlagsmengen bzw. ggf. auch Niederschlagsdauern pro Zeitrahmen zu tun.

Im Endergebnis haben wir vier verschiedene Variablen in unterschiedlichen Verdichtungen (Stunde, Tag, Woche, Monat und Jahreszeit)
Es könnte ggf. nochmal Sinn machen diese in zwei Paaren zu untersuchen: Niederschlagsdauer vs. Menge (die Frage/Hypothese dahinter wäre: gibt es Monate in denen es lage aber nicht heftig regnet vs. solche wo 
es nur zu kurzen aber heftigen Regengüssen kommt?) und die Frage nach der Paarung Luftfeuchtigkeit und Temperatur (feuchte vs. trockene Hitze/Kälte macht einen großen Unterschied beim subjektiven
Temperaturempfinden)

In einer finalen Visualisierung lassen sie diese Erkenntnisse ggf. nochmals prägnanter Visualisieren (eine Heatmap comes to mind).


- can we find historic KPIs to compare our findings (eg. average temperature in January in Kiel 1900)

What Visualisations make sense for our kind of data/insights? Research and try&error

## Wrap up - outlook date-time / time-series
what potential problems have not been adressed?

## Wrap up - What's next/out there?
Insights from data / data vis

What potential problems have not been adressed? (Time series package? etc?)
